Multi-level modeling with Openflexo/\FML

### 2022-03-15 10-conclusions.tex
Synchro après réunion du 15 mars

We fulfill all the challenge requirements and we propose a tooling that
demonstrates the usability of our solution. We developed an ad hoc solution,
including models and metamodels, thanks to the flexibility provided by the
metametamodeling infrastructure offered by Openflexo. This infrastructure
helped us to overcome the limitations of the strict modeling paradigm mentioned
in Sect.~\ref{sec:introduction}, this is, the lack of support for different
forms of classifications and for the duality type/object. In that sense, our
solution seamlessly integrates the type-object pattern which allows us to
dynamically create different \emph{instances} of model elements (\eg, \emph{TaskType}, among others) and use them for typing other model elements. This, together with the \FML core support for (level agnostic) classical linguistic instantiation, permitted us to merge linguistic and ontological instantiation in order to provide a satisfactory solution to the Multi-Level Process challenge.

The features of \FML enable a great modeling flexibility which, coupled with an appropriate methodology, allow us to achieve 
the multi-level capabilities without dedicated tooling, in order to solve
problems which require it. We use virtual models to \enquote{implement} levels on demand.

>>>>

We fulfill all the challenge requirements and we propose a tooling that
demonstrates the usability of our solution. We developed an ad hoc solution,
including models and metamodels, thanks to the flexibility provided by the
metametamodeling infrastructure offered by Openflexo. This infrastructure
helped us to overcome the limitations of the strict modeling paradigm mentioned
in Sect.~\ref{sec:introduction}, this is, the lack of support for different
forms of classifications and for the duality type/object. In that sense, our
solution seamlessly integrates the type-object pattern which allows us to
dynamically create different \emph{instances} of model elements (\eg, \emph{TaskType}, among others) and use them for typing other model elements. This, together with the \FML core support for (level-agnostic) classical linguistic instantiation, permitted us to merge linguistic and ontological instantiation in order to provide a satisfactory solution to the Multi-Level Process challenge.

The features of \FML enable modeling flexibility which, coupled with an appropriate methodology, allow us to achieve 
the multi-level capabilities without dedicated tooling, in order to solve
problems which require it. We use virtual models to \enquote{implement} levels on demand.

### 2022-03-15 6-modelpresentation.tex
Synchro après réunion du 15 mars

The presentation of our solution to the \mlpc follows a systematic methodology where modeling choices are introduced step by step, as the requirements are stated. The satisfaction of requirements is explained throughout this presentation.

%\todo[inline]{phrase d'intro pour expliquer que les req étaient couverts au fur et à mesure (côté systématique)}

The designed multilevel architecture shown in Fig.~\ref{fig:MultilevelArchitecture} captures the two use cases described in the Process Challenge.

Note that in all following figures, blue is used to represent \FML virtual
models and concepts (\textsf{VirtualModel} and \textsf{FlexoConcept} in
Fig.~\ref{fig:mm}), see
Fig.~\ref{fig:BPMNSubsetExample},~\ref{fig:MultilevelArchitecture},~\ref{fig:LinguisticAndOntologicInstantiation},~\ref{fig:BaseMetamodel},~\ref{fig:AcmeFullArchitecture},~\ref{fig:ToolingArchitecture}, while brown is used for instances of
\textsf{FlexoConcept}, see Fig.~\ref{fig:BPMNSubsetExample},~\ref{fig:MultilevelArchitecture},~\ref{fig:LinguisticAndOntologicInstantiation},~\ref{fig:AcmeFullArchitecture},~\ref{fig:ToolingArchitecture}.
% 3, 4, 5, 6, 7, 8, 9
% Expliquer que le use case XSure ne nécessitait pas l'extension de metamodèle requise pour le usecase Acme Software Developement Process

@@

Our proposition relies on ontological instantiation as presented in Fig.~\ref{fig:LinguisticAndOntologicInstantiation}, with a common root concept \texttt{ModelingElement}. Two types of ontological instantiation are needed to meet the requirements of the challenge. One provides that some instances conform to only one type (\texttt{Process} and \texttt{Task} do respectively conform to \texttt{ProcessType} and \texttt{TaskType}). While some entities define their conformity to several types (\texttt{Actor} and \texttt{Artifact} do respectively conform to several \texttt{ActorType} and \texttt{ArtifactType}). These instantiations are respectively expressed using the relations \texttt{type} and \texttt{types} between \texttt{Type}, \texttt{Instance} and \texttt{MultiInstance} concepts. In the diagram, the realization of these relations are indicated by a derived relation, \texttt{/type} or \texttt{/types}, as for example between \texttt{Process} and \texttt{ProcessType} or between \texttt{Actor} and \texttt{ActorType}.

@@

We first present the process definition part of the base metamodel, located on the left of Fig.~\ref{fig:BaseMetamodel}.

@@

The \texttt{ActorType} concept inherits from the \texttt{Type} concept and the \texttt{allowedActorTypes} relation to \texttt{ActorType} defined in \texttt{TaskType} (with 0..* cardinality) captures \textbf{P5} requirement. Requirement \textbf{P6} is symmetrically satisfied with \texttt{allowedActors} relation to \texttt{Actor} also defined in \texttt{TaskType} (with 0..* cardinality). The same modeling pattern applies to \texttt{ArtefactType} also extending the concept \texttt{Type}, and both relations \texttt{usedArtifactTypes} and \texttt{producedArtifactTypes} defined in \texttt{TaskType} (\textbf{P7}). \texttt{TaskType} additionally exposes an \texttt{expectedDuration} role (expressed in number of days), satisfying \textbf{P8}.

The \texttt{Actor} concept defines a boolean attribute called \texttt{isSenior}, while \texttt{TaskType} defines an additional \texttt{isCritical} boolean attribute, indicating that some instances are flagged as critical and must be performed by senior actors. To fulfill \textbf{P9} requirement, a supplementary constraint is required for \texttt{TaskType} and is captured through the following invariant expressed in the \FML language:

@@

This invariant should be completed with additional constraints defined in the \texttt{Task} concept, which apply to performing actors assigned to enact tasks.

\subsubsection{Process enactment}
\label{sec:ProcessEnactment}
We now present the process enactment part of base metamodel, located right of Fig.~\ref{fig:BaseMetamodel}. All concepts defined in this subsection are either specializations of the \texttt{Instance} concept (if they have exactly one type) or the \texttt{MultiInstance} concept (when they have several types).

@@

The semantics is unclear relatively to the instantiation policy for artifacts.
In the requirements, nothing is explicitly stated about the link between the
\emph{artifact type} of a \emph{task type} and the types of the artifacts of a
corresponding task.  Here, we assume that task execution implies that for each
used and produced \emph{artifact type} defined in a related \emph{task type},
it exists at least one artifact of the right type. The following excerpt of
\FML code shows a partial implementation of this. We proceed likewise for
produced artifacts.

@@

Each concept inherits from \texttt{ModelingElement}, which defines a \texttt{lastUpdated} attribute with \texttt{Date} type, and thus satisfies \textbf{P19} requirement.

@@

The challenge describes in a second part a Software engineering process for a fictional Acme company. The Base metamodel described previously is too generic to capture all domain-specific aspects of this use case. We chose to complete the architectural hierarchy with a specific virtual model, specific to the Acme software development process metamodel, shown in Fig.~\ref{fig:AcmeFullArchitecture}.
The \textit{Acme} metamodel specializes the Base metamodel, and \textit{Acme model} is an instance of the \textit{Acme} metamodel. The metamodel inheritance is implemented by virtual model's inheritance. Any instance of the Acme model is either an instance of a concept defined in the Base metamodel, or a concept defined in the specialized \textit{Acme} metamodel.
% FML follows a classical object-oriented semantics. ???

@@

The Fig.~\ref{fig:AcmeFullArchitecture} highlights all conceptual levels of the architecture. This figure is only partial at instance level and not all instances are represented. It also contains an instance of enacted a \textit{Software Engineering Process} at the bottom. The Acme metamodel specializes some concepts of the Base metamodel, for example, \texttt{SETaskType} extends \texttt{TaskType} and \texttt{SEArtifactType} extends \texttt{ArtifactType}. These concepts are further specialized  \texttt{CodingTaskType} extends \texttt{SETaskType} and \texttt{CodeArtifactType} extends \texttt{SEArtifactType}.
The Acme metamodel also defines a specialized task type \texttt{Coding} extending \texttt{Task}, and provides the \texttt{Developer} concept extending \texttt{ActorType}. This metamodel is completed with the \texttt{ProgrammingLanguage} concept, defined as an enumeration (\texttt{Java}, \texttt{C}, \texttt{COBOL}). All the instances required to capture the challenge use case are defined in the final Acme model (lower part of the figure). Instances are represented with rounded boxes and linguistic instantiation is represented using dashed connectors.

@@

The capture of \textbf{S3} requirement is performed while defining \inst{Coding} as an instance of concept \texttt{CodingTaskType} and defining a relation \texttt{languages} to \texttt{ProgrammingLanguage} with 1..* cardinality.
The \texttt{newTask(String)} behavior overrides the generic implementation by instantiating a \texttt{CodingTask} concept, specializing \texttt{Task}, as shown in the following excerpt of \FML code:

@@

\textbf{S5} is more ambiguous as a task type \texttt{CodingTask} defines one or more programming languages but produces code which is expressed in one language only. This requirement is captured through the redefinition of the behavior \texttt{declaresRequiredProducedArtifacts} where programming language should also match.
\textbf{S6} is guaranteed though the \texttt{languages} relation defined in \texttt{CodeArtifact} and the following invariant declared in \texttt{CodeArtifact}:
\begin{lstlisting}
forEach (artifactType : types) {
  assert !(artifactType instanceof CodeArtifactType) | artifactType.doesImplement(languages)
}
\end{lstlisting}

@@

\textbf{S13} requirement has been previously partially fulfilled. This must be completed with an association between an artifact produced and the task that validates it. This is modeled by the reference to \inst{Test Cases} as a produced artifact type in \inst{Test Case design} and the reference to \inst{Test Cases} as a used artifact in \inst{Test Design review}, as shown in Fig.~\ref{fig:AcmeSoftwareDevelopmentProcess}.

@@

The other tool, called \textit{Enacted process graphical editor} and shown on Fig.~\ref{fig:ScreenshotEnactedProcessEditor}, provides process edition and tasks assignations through a graphical visualization displaying the process being executed. This tool is represented on the right side of the Fig.~\ref{fig:ToolingArchitecture}. It follows the same architecture pattern of the \textit{ProcessType graphical editor}, with two model slots referencing both the model and a diagram. Process enactment is operated from a \texttt{ProcessType}, and must be defined using a name identifying the newly instantiated process. All tasks are created from their \texttt{TaskType} definition, and required assignments apply (if for example \inst{Coding} \texttt{TaskType} defines COBOL as output programming language, the related task is automatically assigned to \inst{Ann Smith}). Each task has a status as well as a set of performing actors, a begin and end dates, and used and produced artifacts.

>>>>

The presentation of our solution to the \mlpc follows a systematic methodology where modeling choices are introduced step by step, as the requirements are stated. The satisfaction of requirements is explained throughout this presentation.
%\todo[inline]{phrase d'intro pour expliquer que les req étaient couverts au fur et à mesure (côté systématique)}

The designed multilevel architecture shown in Fig.~\ref{fig:MultilevelArchitecture} captures the two use cases described in the Process Challenge.
Note that in all following figures, blue is used to represent \FML virtual
models and concepts (\textsf{VirtualModel} and \textsf{FlexoConcept} in
Fig.~\ref{fig:mm}), see
Fig.~\ref{fig:BPMNSubsetExample},~\ref{fig:MultilevelArchitecture},~\ref{fig:LinguisticAndOntologicInstantiation},~\ref{fig:BaseMetamodel},~\ref{fig:AcmeFullArchitecture},~\ref{fig:ToolingArchitecture}, while brown is used for instances of
\textsf{FlexoConcept}, see Fig.~\ref{fig:BPMNSubsetExample},~\ref{fig:MultilevelArchitecture},~\ref{fig:LinguisticAndOntologicInstantiation},~\ref{fig:AcmeFullArchitecture},~\ref{fig:ToolingArchitecture}.{\color{red} color scheming is good, but fails when document printed in gray scale.  It might be worthwhile to consider patterns as well.}
% 3, 4, 5, 6, 7, 8, 9
% Expliquer que le use case XSure ne nécessitait pas l'extension de metamodèle requise pour le usecase Acme Software Developement Process

@@

Our proposition relies on ontological instantiation as presented in Fig.~\ref{fig:LinguisticAndOntologicInstantiation}, with a common root concept \texttt{ModelingElement}. Two types of ontological instantiation are needed to meet the requirements of the challenge. One provides that some instances conform to only one type (\texttt{Process} and \texttt{Task} do respectively conform to \texttt{ProcessType} and \texttt{TaskType}) {\color{red}Rephrase the previous sentence.  "One provides ..."  where do you describe the second?}. While some entities define their conformity to several types (\texttt{Actor} and \texttt{Artifact} do respectively conform to several \texttt{ActorType} and \texttt{ArtifactType}). These instantiations are respectively expressed using the relations \texttt{type} and \texttt{types} between \texttt{Type}, \texttt{Instance} and \texttt{MultiInstance} concepts. In the diagram, the realization of these relations are indicated by a derived relation, \texttt{/type} or \texttt{/types}, as for example between \texttt{Process} and \texttt{ProcessType} or between \texttt{Actor} and \texttt{ActorType}.

@@

We first{\color{red} Use "first" when you have second third coming up later} present the process definition part of the base metamodel, located on the left of Fig.~\ref{fig:BaseMetamodel}.

@@

The \texttt{ActorType} concept inherits from the \texttt{Type} concept and the \texttt{allowedActorTypes} relation to \texttt{ActorType} defined in \texttt{TaskType} (with 0..* cardinality) captures the requirement \textbf{P5}. Requirement \textbf{P6} is symmetrically satisfied with \texttt{allowedActors} relation to \texttt{Actor} also defined in \texttt{TaskType} (with 0..* cardinality). The same modeling pattern applies to \texttt{ArtefactType} also extending the concept \texttt{Type}, and both relations \texttt{usedArtifactTypes} and \texttt{producedArtifactTypes} defined in \texttt{TaskType} (\textbf{P7}). \texttt{TaskType} additionally exposes an \texttt{expectedDuration} role (expressed in number of days), satisfying \textbf{P8}.

The \texttt{Actor} concept defines a boolean attribute called \texttt{isSenior}, while \texttt{TaskType} defines an additional \texttt{isCritical} boolean attribute, indicating that some instances are flagged as critical and must be performed by senior actors. To fulfill \textbf{P9}, a supplementary constraint is required for \texttt{TaskType} and is captured through the following invariant expressed in the \FML language:

@@

This invariant should be completed with additional constraints defined in the \texttt{Task} concept, which applies to performing actors assigned to enact tasks.

\subsubsection{Process enactment}
\label{sec:ProcessEnactment}
We now present the process enactment part of base metamodel, located at the right of Fig.~\ref{fig:BaseMetamodel}. All concepts defined in this subsection are either specializations of the \texttt{Instance} concept (if they have exactly one type) or the \texttt{MultiInstance} concept (when they have several types).

@@

The semantics is relatively unclear to the instantiation policy for artifacts.
In the requirements, nothing is explicitly stated about the link between the
\emph{artifact type} of a \emph{task type} and the types of the artifacts of a
corresponding task.  Here, we assume that task execution implies that for each
used and produced \emph{artifact type} defined in a related \emph{task type},
it exists at least one artifact of the right type. The following excerpt of
\FML code shows a partial implementation of this. We proceed likewise for
produced artifacts.

@@

Each concept inherits from \texttt{ModelingElement}, which defines a \texttt{lastUpdated} attribute with \texttt{Date} type, and thus satisfies the requirement \textbf{P19}.

@@

The challenge describes a software engineering process for a fictional Acme company in the second part. The Base metamodel described previously is too generic to capture all domain-specific aspects of this use case. We chose to complete the architectural hierarchy with a specific virtual model, specific to the Acme software development process metamodel, shown in Fig.~\ref{fig:AcmeFullArchitecture}.
The \textit{Acme} metamodel specializes the Base metamodel, and \textit{Acme model} is an instance of the \textit{Acme} metamodel. The metamodel inheritance is implemented by virtual model's inheritance. Any instance of the Acme model is either an instance of a concept defined in the Base metamodel, or a concept defined in the specialized \textit{Acme} metamodel.
% FML follows a classical object-oriented semantics. ???

@@

The Fig.~\ref{fig:AcmeFullArchitecture} highlights all conceptual levels of the architecture. This figure is only partial at instance level and not all instances are represented. It also contains an instance of enacted a \textit{Software Engineering Process} at the bottom. The Acme metamodel specializes some concepts of the Base metamodel, for example, \texttt{SETaskType} extends \texttt{TaskType} and \texttt{SEArtifactType} extends \texttt{ArtifactType}. These concepts are further specialized  \texttt{CodingTaskType} extends \texttt{SETaskType} and \texttt{CodeArtifactType} extends \texttt{SEArtifactType}.
The Acme metamodel also defines a specialized task type \texttt{Coding} extending \texttt{Task}, and provides the \texttt{Developer} concept extending \texttt{ActorType}. This metamodel is completed with the \texttt{ProgrammingLanguage} concept, defined as an enumeration (\texttt{Java}, \texttt{C}, \texttt{COBOL}). All the instances required to capture the use case of the challenge are defined in the final Acme model (lower part of the figure {\color{red} missing reference}). Instances are represented with rounded boxes and linguistic instantiation is represented using dashed connectors.

@@

The requirement \textbf{S3} is captured by defining \inst{Coding} as an instance of concept \texttt{CodingTaskType} and defining a relation \texttt{languages} to \texttt{ProgrammingLanguage} with 1..* cardinality.
The \texttt{newTask(String)} behavior overrides the generic implementation by instantiating a \texttt{CodingTask} concept, specializing \texttt{Task}, as shown in the following excerpt of \FML code:

@@

\textbf{S5} is more ambiguous as a task type \texttt{CodingTask} defines one or more programming languages but produces code which is expressed in only one language. This requirement is captured through the redefinition of the behavior \texttt{declaresRequiredProducedArtifacts} where programming language should also match.
\textbf{S6} is guaranteed though the \texttt{languages} relation defined in \texttt{CodeArtifact} and the following invariant declared in \texttt{CodeArtifact}:
\begin{lstlisting}
forEach (artifactType : types) {
  assert !(artifactType instanceof CodeArtifactType) | artifactType.doesImplement(languages)
}
\end{lstlisting}

@@

The requirement \textbf{S13} has been previously partially fulfilled. This must be completed with an association between an artifact produced and the task that validates it. This is modeled by the reference to \inst{Test Cases} as a produced artifact type in \inst{Test Case design} and the reference to \inst{Test Cases} as a used artifact in \inst{Test Design review}, as shown in Fig.~\ref{fig:AcmeSoftwareDevelopmentProcess}.

@@

The other tool, called \textit{Enacted process graphical editor} and shown on Fig.~\ref{fig:ScreenshotEnactedProcessEditor}, provides process edition and tasks assignations through a graphical visualization displaying the process being executed. This tool is represented on the right side of the Fig.~\ref{fig:ToolingArchitecture}. It follows the same architecture pattern of the \textit{ProcessType graphical editor}, with two model slots referencing both the model and a diagram. Process enactment is operated from a \texttt{ProcessType}, and must be defined using a name identifying the newly instantiated process. All tasks are created from their \texttt{TaskType} definitions, and required assignments apply (if for example \inst{Coding} \texttt{TaskType} defines COBOL as output programming language, the related task is automatically assigned to \inst{Ann Smith}). Each task has a status as well as a set of performing actors, a begin and end dates, and used and produced artifacts.

### 2022-03-15 7-requirementssatisfaction.tex
Synchro après réunion du 15 mars

    \item \emph{Linguistic instantiation} (\textit{L.Inst.}): a requirement is satisfied by the instantiation of one or more concepts (for example for \textbf{S1}: \inst{Requirement analysis} is defined as an instance of \texttt{SETaskType}).

>>>>

    \item \emph{Linguistic instantiation} (\textit{L.Inst.}): the requirements are satisfied by the instantiation of one or more concepts (for example for \textbf{S1}: \inst{Requirement analysis} is defined as an instance of \texttt{SETaskType}).

### 2022-03-15 8-discussion.tex
Synchro après réunion du 15 mars

  Due to the fact that our approach is level-agnostic, our solution could have
  more or fewer levels depending on the variations of the use case. However,
  the number of levels is related to the problem, our solution uses 3 levels.

@@

  Thanks to the level-agnosticism nature of the Openflexo approach, cross-level
  relationships are not an issue. Model elements of different levels can be
  linked each other in a transparent way, using inheritance or instantiation.
  For instance, in the Fig.~\ref{fig:AcmeFullArchitecture}, the
  \texttt{Developer} concept from \texttt{Acme} metamodel specializes the
  \texttt{ActorType} of the base metamodel. The \texttt{Software Engineering Process} of the
  \texttt{Acme} model is an instance of \texttt{ProcessType} in the metamodel.
%  \textcolor{red}{Phrase faite a relire sur Developer\dots}

@@

  As for cross-level relationships, cross-level constraints do not have a
  specific nature. They are like any constraint a user can define by adding a
  behavior to a model element. Therefore, if a constraint is mandatory when
  establishing a relationship between elements from different levels, the user
  has to create it manually. This can be a limit of our approach: the cost of
  the flexibility of our tooling is a limited number of automated behaviors.

@@

  A strength of the  model federation approach we have adopted resides in its
  flexibility. As we build metamodels and models together instead of fitting
  into a metamodel, our solution is more flexible. Therefore one can extend a
  solution easily. The challenge itself can be seen as a validation of the
  extensibility ability of our solution: it was decomposed into two steps
  (Xsure process, then ACME process) which can be seen as a simulation of the
  evolution of the  specification. We observed that our approach has been
  resilient to change.

>>>>

  Due to the fact that our approach is level-agnostic, our solution could have
  more or fewer levels depending on the variations of the use case. However,
  the number of levels is related to the problem.  For example, our solution uses 3 levels for this multi-level process challenge.

@@

  Thanks to the level-agnosticism nature of the Openflexo approach, cross-level
  relationships are not an issue. Model elements of different levels can be
  linked to each other in a transparent way, using inheritance or instantiation.
  For instance, in the Fig.~\ref{fig:AcmeFullArchitecture}, the
  \texttt{Developer} concept from \texttt{Acme} metamodel specializes the
  \texttt{ActorType} of the base metamodel. The \texttt{Software Engineering Process} of the
  \texttt{Acme} model is an instance of \texttt{ProcessType} in the metamodel.
%  \textcolor{red}{Phrase faite a relire sur Developer\dots}

@@

  As for cross-level relationships, cross-level constraints do not have a
  specific nature. They are like any constraint that a user can define by adding a
  behavior to a model element. Therefore, if a constraint is mandatory when
  establishing a relationship between elements from different levels, the user
  has to create it manually. This can be a limit of our approach; the cost of
  the flexibility of our tooling is a limited number of automated behaviors.

@@

  A strength of the  model federation approach we have adopted resides in its
  flexibility. As we build metamodels and models together instead of fitting
  into a metamodel, our solution is more flexible. Therefore one can extend a
  solution easily. The challenge itself can be seen as a validation of the
  extensibility of our solution: it was decomposed into two steps
  (Xsure process, then ACME process) which can be seen as a simulation of the
  evolution of the  specification. We observed that our approach has been
  resilient to change.

### 2022-03-15 9-relatedworks.tex
Synchro après réunion du 15 mars

A plethora of multi-level modeling approaches and tools\footnote{\url{https://homepages.ecs.vuw.ac.nz/Groups/MultiLevelModeling/MultiTools}} with different foundations have appeared in recent years~\parencite{somogyi2021playground}. Comparing them lies out of the scope of this paper. In this sense, we limit the following discussion to the presentation and comparison of previous solutions to the MULTI Process Challenge.

@@

More similar to us,~\textcite{deeptelos2019} uses DeepTelos
\parencite{deeptelos2016}, an extension of the Telos language
\parencite{telos1990} in order to solve the process challenge. As with
Openflexo/\FML, they do not use explicit (numbered) levels nor potency.
However, unlike Openflexo, DeepTelos integrates a multi-level modeling specific
construct similar to the PowerType \parencite{atkinson2001essence} pattern they
call \emph{most general instances} which the authors use to simulate potency.
Models in DeepTelos are organized in (tree-like) module hierarchies, where
submodules can see all the concepts defined in parent modules. The authors use
this module system in order to \emph{organize} their solution for the
Multi-level process challenge. Concretely they created a hierarchy of modules
such as the top module contains base concepts and formulas required for
multi-level modeling (\eg, the support for \emph{most general instances}), and a sub-module contains process definitions fulfilling requirements P1 to P19. This sub-module contains in turn a sub-module representing the coding process type whereas concrete coding processes are represented in subsequent sub-modules. There are two main differences between the DeepTelos and Openflexo/\FML: 1) we use the type-object pattern instead of the powertype pattern; 2) DeepTelos reifies their support for multi-level modeling while our remains ad hoc.

>>>>

A plethora of multi-level modeling approaches and tools\footnote{\url{https://homepages.ecs.vuw.ac.nz/Groups/MultiLevelModeling/MultiTools}} with different foundations have appeared in recent years~\parencite{somogyi2021playground}. A complete comparison of all these approaches is out of the scope of this paper. In this sense, we limit the following discussion to the presentation and comparison of previous solutions to the MULTI Process Challenge.

@@

More similar to us,~\textcite{deeptelos2019} uses DeepTelos
\parencite{deeptelos2016}, an extension of the Telos language
\parencite{telos1990} in order to solve the process challenge. As with
Openflexo/\FML, they do not use explicit (numbered) levels nor potency.
However, unlike Openflexo, DeepTelos integrates a multi-level modeling specific
construct similar to the PowerType \parencite{atkinson2001essence} pattern they
call \emph{most general instances} which the authors use to simulate potency.
Models in DeepTelos are organized in (tree-like) module hierarchies, where
submodules can see all the concepts defined in parent modules. The authors use
this module system in order to \emph{organize} their solution for the
Multi-level process challenge. Concretely they created a hierarchy of modules
such as the top module contains base concepts and formulas required for
multi-level modeling (\eg, the support for \emph{most general instances}), and a sub-module contains process definitions fulfilling requirements P1 to P19. This sub-module in turn contains another sub-module representing the coding process type whereas concrete coding processes are represented in subsequent sub-modules. There are two main differences between the DeepTelos and Openflexo/\FML: 1) we use the type-object pattern instead of the powertype pattern; 2) DeepTelos reifies their support for multi-level modeling while our remains ad hoc.
